---
title: "Increased Female Employment Rates and Lower Birth Rates"
subtitle: "An Analysis of Canada's Female Employment Rate and Birth Rate"
author: Xu Qi
thanks: "Code and data are available at: https://github.com/xuqi2002/female-employment-and-birth-rate.git"
date: today
date-format: long
abstract: "This paper utilizes data from The World Bank to examine the relationship between Canada's female employment rate and birth rate between 2009 to 2019. The analysis shows a negative correlation between the two variables, indicating that as the female employment rate increased, the birth rate falls. This finding is important because it highlights the importance of work-life balance to national fertility rates. These insights can guide the government to formulate appropriate childcare policies, thereby optimizing the country's demographic structure."
format: pdf
number-sections: true
bibliography: references.bib
---

# Introduction

The relationship between a country's female employment rate and birth rate is a topic that has attracted considerable attention and debate among economists, population scientists and policymakers. The relationship is complex and multifaceted, often reflecting broader social and economic dynamics, cultural norms and policy environments. The female employment rate, defined as the proportion of the female population aged 15 and older that is employed, is an indicator of gender equality and economic participation [@The_World_Bank_2024a]. As a developed country with a diversified economy and progressive social policies, the study of this relationship in Canada can shed light on broader discussions about work-life balance, gender equality and demographic trends. Over the past several decades, Canada has seen a significant increase in female labor force participation, mirroring trends in other advanced economies. This shift has been attributed to a variety of factors, including increased educational attainment among women, changes in societal attitudes towards female employment, and the implementation of policies aimed at facilitating work-life balance, such as maternity leave and childcare support [@Blau_Kahn_2017]. 

At the same time, birth rates have been declining in many developed countries, including Canada. The birth rate, measured as the number of live births per 1,000 population per year, is influenced by myriad factors, including economic conditions, access to contraception and reproductive health services, and personal choices regarding family size and childbearing timing [@Sobotka_2017]. In the case of Canada, the birth rate has experienced fluctuations over the past few decades, with recent trends showing a gradual decline. This decline in birth rates is a feature common to many advanced economies and is often associated with factors such as increased female educational attainment and labor force participation, urbanization, and changes in family planning preferences [@Sobotka_2017].

In this paper, we will examine the relationship between Canada's female employment rate and birth rate through a linear regression analysis. The estimand here is how female employment rate and birth rate are related. We will download data from the Data Bank website of The World Bank [@The_World_Bank_2024b]. Our respondents of interest are the percentage of the female working-age population, as they represent the potential labor force and have a significant impact on a country's economic growth and development. Based on the analysis, we found that there is a negative relationship between female employment rate and birth rate.

In the data section, we discuss the source of data used in this paper, the strengths and weaknesses of The World Bank, methodologies that follow it, and data terminology. Then we present the data visualization, focusing on the trajectory of female employment rate and birth rate from 2009 to 2019 in Canada. In the model section, we will analyze the trend by creating a linear regression model. In the result section, we will present the result of the model in a graph. Finally, we explore the factors that contribute to this negative relationship between female employment rate and birth date. We will also examine the weakness, and potential next step of this analysis. 

# Data

## Data Description and Methodology

The data used in this paper is obtained from the Data Bank website of The World Bank, it is public data which available through the Data Bank website [@The_World_Bank_2024b]. The World Bank's Data Bank is a comprehensive database that offers free and open access to a vast collection of global development data. It encompasses a wide array of statistical indicators and data sets covering various aspects of economic and social development. These include, but are not limited to, population demographics, economic performance, health and education statistics, environmental indicators, and much more.

The data in the World Bank's Data Bank comes from a variety of sources, with a significant portion being collected and compiled by the World Bank itself, as well as from specialized agencies of the United Nations (UN), international organizations, national statistical offices, and other authoritative public and private sector sources. These data are then aggregated and analyzed by the World Bank to identify trends and inform policy recommendations.

The two datasets that I will be using are: `"Birth rate, crude, per 1000 people"` [@The_World_Bank_2024d], and `"Employment to population ratio, 15+, female (%) (national estimate)"` [@The_World_Bank_2024c]. The two original datasets include data from 265 countries around the world for the period 1960-2022, although data are not available for every country in every year. I processed these two datasets, keeping only Canada's data for 2009-2019, and merged them into one dataset for analysis. The birth rate dataset presents the number of live births per 1,000 people in a given population over one year. The observations recorded in the processed dataset was 11.4 in 2009, and 9.9 in 2019. The female employment rate dataset presents the percentage of a country's female population aged 15 years and older that is employed. The value is expressed as a percentage of the total female population within the same age group. The observations recorded in the processed dataset was 57.958 in 2009, and 58.453 in 2019.

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(readr)
library(dplyr)
library(ggplot2)

analysis_data <- read_csv(here::here("data/analysis_data/canada_2009_2019.csv"))

```

```{r}
#| echo: false
#| warning: false
#| label: tbl-data
#| tbl-cap: "A summary table of cleaned data"

library(knitr)

kable(analysis_data)

```
@tbl-data presents the overview of the cleaned dataset. There are 11 observations which are years from 2009 to 2019. The variables in the dataset include `Year`,  `Birth_Rate_per_1000_people` and `Female_Employment_Rate` (in percentage).

In this paper, the analysis is carried out using the statistical programming language `R` [@r], with the help of extra packages including `arrow` [@arrow], `dplyr` [@Dplyr], `ggplot2` [@Ggplot2], `here` [@herePackage], `kableExtra` [@citekableExtra], `knitr` [@knitr], `modelsummary` [@model], `readr` [@readr], `tibble` [@tibblePackage], and `tidyverse` [@rTidyverse].

## Data Visualization

### Trend of Female Employment Rate
```{r fig.width=12, fig.height=8}
#| label: fig-1
#| fig-cap: Canada’s female employment rate from 2009 to 2019. 
#| echo: false
#| warning: false

library(ggplot2)

analysis_data$Year <- factor(analysis_data$Year)

ggplot(analysis_data, aes(x = Year, y = Female_Employment_Rate, group = 1)) +
  geom_line(color = "darkgreen") + 
  geom_point(color = "darkgreen") +
  theme_minimal() +
  scale_x_discrete(name = "Year", breaks = levels(analysis_data$Year)) +
  labs( y = "Female Employment Rate (%)")

```
@fig-1 shows the trend in the female employment rate in Canada from 2009 to 2019. The line graph indicates an overall upward trajectory in the rate of female employment during this decade. The female employment rate was initially 54.616% in 2009. After that, there is a sharp decrease until 2011. From 2011 to 2016, there are some volatility, with the female employment rate experiencing both increases and decreases but generally trending downwards during this period. After 2016, the rate begins to rise again, with reaching the peak in 2019 with 58.453%. This figure suggests that more women are participating in the Canadian labor force over time, though social, economic and policy changes may have influenced the rate's variability throughout these years.

\newpage

### Trend of Birth Rate
```{r fig.width=12, fig.height=8}
#| label: fig-2
#| fig-cap: Canada’s birth rate from 2009 to 2019. 
#| echo: false
#| warning: false

library(ggplot2)

analysis_data$Year <- factor(analysis_data$Year)

ggplot(analysis_data, aes(x = Year, y = Birth_Rate_per_1000_people, group = 1)) +
  geom_line(color = "red") + 
  geom_point(color = "red") +
  theme_minimal() +
  scale_x_discrete(name = "Year", breaks = levels(analysis_data$Year)) +
  labs( y = "Birth Rate (per 1000 people)")

```

@fig-2 presents a line graph showing the trend of Canada's birth rate for a decade from 2009 to 2019, which is a clear downward trend. Starting at 11.4  births per 1,000 people in 2009, the birth rate consistently declines year over year, reaching its lowest point after a decade in 2019, with only 9.9 newborns per 1000 people. Such a descending trend may suggests various social and economic changes, such as shifts in societal norms regarding family size, economic incentives or disincentives for having children, or other demographic changes within the country. The consistent downward trajectory suggests that these factors are persisting over time, contributing to a sustained reduction in the number of births.

\newpage

# Model

After conducting basic analysis on the dataset, we have observed a negative correlation between the female employment rate and the birth rate, indicating a potential linear relationship. To gain further insights, we will be implementing a linear regression model.

Here is the equation of the linear regression model.

$$ 
Yij = β0 + β1Xij + εij
$$

where:

-   $Y$ is the Birth Rate

-   $X$ is the Female Employment Rate

-   $Y_{ij}$ is the Birth Rate for observation j in year i.

-   $X_{ij}$ is the Female Employment Rate for observation j in year i.

-   $\beta_0$ is the intercept/constant term, which represents the expected value of Birth Rate when the Female Employment Rate is equal to zero.

-   $\beta_1$ is the slope coefficient or the estimated change in Birth Rate for a one-unit increase in the Female Employment Rate.

-   $\epsilon_{ij}$ is the error term or the deviation of the actual value of Birth Rate from the predicted value based on the regression equation.

The purpose of a linear regression model in this case is to estimate the values of β0 and β1, so that the model fits the data well, and predicts the expected value of the newborn birth rate for different values of the female employment rate. The statistical significance of β1 can be assessed using a t-test, which tests whether the estimated coefficients are significantly different from zero. If the p-value of the t-test is less than the selected level of significance, we can conclude that there is a significant relationship between the female employment rate and the newborn birth rate.

\newpage

# Result

```{r}
#| label: fig-3
#| fig-cap: "The model shows the relationship between Female Employment Rate and Birth Rate"
#| echo: false
#| warning: false

library(ggplot2)

analysis_data %>%
  ggplot(aes(x = `Female_Employment_Rate`, y = `Birth_Rate_per_1000_people`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  labs(x = "Female Employment Rate (%)", y = "Birth Rate (per 1000 people)") +
  theme_classic()

```

@fig-3 illustrates a fitted graph of the linear regression model, showing the relationship between the Female Employment Rate (on the x-axis) and the Birth Rate per 1,000 people (on the y-axis). The dashed line indicates the trend as estimated by the linear regression model. The grey shaded area around the dashed line represents the confidence interval for the regression line, giving a visual indication of the uncertainty around the estimated relationship. The plot suggests a negative relationship between the two variables: as female employment rate increases, the birth rate rate decreases. The regression line confirms this trend, with a negative slope that is statistically significant. Overall, this figure provide strong evidence of the negative association between female employment rates and birth rates. With the confidence interval indicating the range, we can be reasonably sure that the true regression line lies in this area. 

```{r}
#| label: tbl-model
#| tbl-cap: "T-Value and P-Value of Regression"
#| echo: false
#| warning: false
library(knitr)

library(modelsummary)

loaded_model <- readRDS(here::here("models/linear_model.rds"))

model_summary <- summary(loaded_model)

t_values <- model_summary$coefficients[, "t value"]
p_values <- model_summary$coefficients[, "Pr(>|t|)"]

summary_df <- data.frame(t_value = t_values, p_value = p_values)
kable(summary_df)
```


```{r}
#| label: tbl-model2
#| tbl-cap: "Summary of the linear regression model"
#| echo: false
#| warning: false

modelsummary::modelsummary(
  list("Female_Employment_Rate" = loaded_model),
  fmt = 2
)
```

@tbl-model and @tbl-model2 are summaries of the linear regression model, they provide several key pieces of information about the relationship between the female employment rate and the birth rate:

* The intercept is approximately 54.01, suggesting that when the female employment rate is 0, the birth rate is predicted to be around 54 births per 1000 people. However, this interpretation of the intercept may not be meaningful if a female employment rate of 0 is not feasible or outside the range of the data.

* The coefficient for Female Employment Rate is approximately -0.75, with a standard error of 0.52. This suggests that for each one percent increase in the female employment rate, the birth rate is expected to decrease by 0.75 births per 1000 people. 

* The R-squared value is 0.186, meaning that approximately 18.6% of the variability in the birth rate is explained by the model. This is a relatively low value, indicating a weak fit of the model to the data.

* The t-value for the Female Employment Rate coefficient is approximately -1.43, indicating the estimate is -1.43 standard deviations away from 0. The negative sign shows that the relationship is negative.

* The p-value for the intercept is approximately 0.1076, suggesting that the intercept is not statistically significant at the common alpha levels 

In summary, this linear model suggests a negative relationship between the Female Employment Rate and the Birth Rate, but neither the intercept nor the slope is statistically significant. Given the low R-squared value, this model also does not appear to be a particularly strong predictor of the birth rate based on the female employment rate. This implies that while the model estimates a negative effect of the increased Female Employment Rate on Birth rate, we do not have sufficient evidence to claim a statistically significant relationship from this sample of data. 

# Discussion

## Evolution of Female Employment in Canada
As we have analyzed above, the trend of the female employment rate in Canada has been marked by significant growth, with fluctuations.This reflect broader economic cycles and policy changes. Over the past few decades, Canada has witnessed substantial changes in the labor market, particularly concerning female labor force participation.

In the early 1990s, Canada's female employment rate began to rise sharply, reflecting a trend towards greater gender equality and the increased need for dual-income households in the face of economic pressures [@Fortin_2005]. This upward trajectory continued into the 2000s, albeit with periodic fluctuations that can be linked to macroeconomic conditions such as recessions and recoveries. For instance, during the economic downturn in 2008-2009, Canada, like many other countries, experienced a decrease in employment rates. However, it is notable that the impact on female employment was less pronounced than on male employment, which some scholars attribute to the concentration of women in less cyclical sectors such as healthcare and education [@Goldin_2014].

Following the recovery from the recession, female employment rates in Canada resumed an upward trend. The participation of women in the workforce has been supported by several factors, including higher educational attainment among women and government policies aimed at encouraging female labor participation, such as subsidized child care and parental leave policies [@Blau_Kahn_2017]. These policies have not only enabled more women to join and remain in the workforce but also to progress into more skilled and higher-paid roles. 

Despite these gains, the trend has not been without its challenges. Women, particularly those with young children, still face significant barriers to full-time employment, leading to a persistent gender gap in full-time employment rates [@Morissette_Ostrovsky_2005]. The part-time work, often undertaken by women, has implications for wage equity and career advancement opportunities. Moreover, the rise in female employment has not been uniform across all sectors or demographics; for example, immigrant women and women in rural areas often have lower employment rates than their urban counterparts [@Hou_Coulombe_2010].

In conclusion, the trend of Canada's female employment rate reflects a complex interplay of economic forces, social change, and policy interventions. While significant progress has been made, ongoing challenges remain, necessitating a continuous commitment to fostering a labor market that supports gender equality and enables all Canadians to contribute to and benefit from economic growth.

## Evolution of Birth Rate in Canada
As we have analyzed above, Canada's birth rate has a declining trend. It is not just happened in the past decade but over several decades, a pattern consistent with many other developed countries. The decline in birth rates reflects significant changes in individual and societal choices and is shaped by the broader social and economic context.

Since the late 1960s, the birth rate of Canada has been on a downward slope, a trend described by the demographic transition theory, which links reductions in birth and death rates to societal advancements and improved living standards [@Kirk_1996]. This transition often accompanies increased urbanization, greater access to education, and improved healthcare, particularly in women's reproductive health services. By the 1990s, the birth rate in Canada had fallen to below-replacement levels, which is the rate required to replace the population without immigration [@Milan_2018]. The factors contributing to the decline are manifold. Increasing female participation in higher education and the workforce has been associated with delayed childbearing and reduced family size [@Beaujot_2007]. The rising cost of living, particularly in urban centers where most of the population resides, and concerns about work-life balance have also led many individuals and couples to opt for fewer children.

Another factor influencing the birth rate is the shifting pattern of immigration. Immigrants tend to have higher birth rates upon arrival in Canada, but these rates often converge to the national average within a generation [@Ng_Nault_1997]. This demographic effect adds complexity to the analysis of birth rate trends. Moreover, social attitudes towards marriage and parenthood have evolved. With increased acceptance of cohabitation and non-traditional family structures, the link between marriage and childbearing has weakened, further affecting birth rates [@Le_Bourdais_Lapierre_Adamcyk_2004]. Government policies also play a role in shaping birth rates. In Canada and other developed countries, policies such as subsidies for child and parental leave provisions aim to alleviate the financial burden of child-rearing, potentially influencing decisions around having children. 

Looking for the future, the continued decline in birth rates will have significant implications for Canada's demographic structure, including an aging population and the reated economic problems.  Policymakers must aviod these implications, considering the adequacy of retirement systems, healthcare services, and labor force planning.

## Impact of Female Employment on Birth Rate

Although the regression analysis in this paper does not strongly support a negative correlation between the female employment rate and birth rate, this relationship has been agreed by many demographers, economists and policymakers, especially as many developed countries are experiencing changes in fertility patterns and the structure of the labor force. 

Economic theory often posits that increased female labor force participation leads to a decrease in birth rates. This inverse relationship is primarily attributed to the opportunity costs associated with childbearing and childrearing. When women engage in paid work, the time and economic resources devoted to having children come at the expense of their career advancement and income potential [@Becker_2009]. Moreover, as women attain higher levels of education and more demanding careers, they may choose to delay marriage and parenthood or opt for smaller family sizes [@Bloom_2003].

From a sociological standpoint, the changing norms and values around gender roles and family structures influence both female employment and fertility decisions. The rise in gender equality movements has contributed to a shift in societal expectations, leading to increased acceptance and support for women pursuing careers alongside or instead of traditional roles as primary caregivers [@McDonald_2000]. Consequently, as women's roles have diversified beyond the domestic sphere, their contributions to household income have become more vital, which can defer or reduce family planning intentions.

However, the relationship is not always like above. The nature of the job market and the availability of family-friendly policies also play significant roles. Countries with supportive policies, such as subsidized childcare and flexible work arrangements, tend to reduce the negative impact of female employment on birth rates. For example, in Canada and some Scandinavian countries, despite high female employment rates, birth rates have not declined as steeply as might be expected, largely due to such supportive policies [@Rindfuss_Guilkey_Morgan_Kravdal_Guzzo_2007].

In conclusion, while it's clear that female employment rates do have an impact on birth rates, the nature of this impact is complex. As societies evolve, the balancing act between employment and childbearing continues to be evolve in response to changing economic conditions, social norms, and policy frameworks.

## Weakness and Next Step

This analysis examining the impact of the female employment rate on the birth rate in Canada and provides valuable insights. However, it has limitations that merit consideration for future research. 

One key weakness is having the bias of omitting other variables. The model currently includes only the female employment rate as a predictor, possibly neglecting other factors that can influence the birth rate. Variables such as average age at marriage, access to contraception, cultural norms, economic conditions, immigration patterns, and government family policies could also significantly impact fertility rates. Ignoring these factors could mislead the effect of female employment on the birth rate.

Another limitation is the assumption of a linear relationship between the two variables. The relationship may be more complex and could not captured by a simple linear model. For example, there could be threshold effects where the impact of female employment on the birth rate changes after reaching a certain level of employment rate. In addition, the data set used only 11 observations, which means 11 years. A larger dataset could provide a more detailed understanding of the relationship. Also, a comparative analysis including other countries could help contextualize the findings within a global framework. 

For next steps, the analysis could be expanded to include other aspects that have been identified in the text above. This would allow for a more comprehensive model that can control complicated factors. Using a multivariate regression analysis could provide deeper insights. 

\newpage
# References